On the Functionality of the Multilayer Perceptron

### 2020-10-03 Thesis/Drafts/Buell-PHYS799-ThesisDraft0.tex
Updates to Written Thesis

@@

@@

\subsection{The Structure of a Neural Network}
\label{subsec-NetworkStructre}

\begin{algorithmic}
\REQUIRE Network with $L$ layers - $0,1,2,....,L-2,L-1$
\REQUIRE Set of layer functions $f^{(l)}, l \in 0,1,2,....,L-2,L-1$ made up of 
\REQUIRE Input sample - $x^{(0)}$

$Z = \{\} \leftarrow$ array to hold pre-activations \\
$X = \{\} \leftarrow$ array to hold final activations \\

\FOR {$l = 1,2,...,L-2,L-1$}

	\item Break layer function into two linear transformation and non-linear activation

	\item Store values for later use in optimizer
	\item $Z \leftarrow Z.add(z^{(l)})$
	\item $X \leftarrow X.add(x^{(l)})$
\ENDFOR

\paragraph*{}As stated previously, a neural network is composed of a series of functions that are called successively to transform features (an input) into a prediction (an output). As shown in eqn. (\ref{eqn-altLayerFunction}), each function feeds directly into the next as to form a sort of computational graph \cite{Goodfellow}.

\paragraph*{}Typically, a layer function can be divided into two portions: (i.) a Linear transformation, with a bias addition, and (ii.) an element-wise activation transformation. This activation function typically is a non-linear function which allows for the modeling of increasingly complex decision-boundaries. Step (i.) is usually in the form of a matrix-vector equation:
\begin{equation}
\label{eqn-LinearTransform}
z^{(l)} = W^{(l)} x^{(l-1)} + b^{(l)}
\end{equation}
Where $W^{(l)}$ is the \textit{weighting-matrix} for layer $l$, $b^{(l)}$ is the \textit{bias-vector} for layer $l$, $z^{(l)}$ are the \textit{linear-activations} for layer $l$ and $x^{(l-1)}$ is the final activations for layer $l-1$.
Step (ii.) is usually given by some \textit{activation function}:
\begin{equation}
\label{eqn-elementActivation}
x^{(l)} = \sigma^{(l)} \big[ z^{(l)} \big]
\end{equation}
Where $x^{(l)}$ is the final activations for layer $l$ and $z^{(l)}$ is given in equation (\ref{eqn-LinearTransform}). $\sigma^{(l)}$ is some activation function, which is often use to enable the modeling of more complex-decision boundaries.

\paragraph*{}In the case of a 2D input array, $x^{(l-1)} \in \mathbb{R}^{(N \times M)}$, the convolutions filters "step" through the input data and repeatedly compute the element-wise product of each $m \times n$ weighting kernel, and the local activation of the $x^{(l-1)}$ array. The step size in each of the 2-dimension is known as the \textit{stride}. Each of the $K$ filters is used to generate a $m \times n$ feature map from the input activation. For an appropriately trained network, some filters may be dedicated to detecting vertical lines, horizontal lines, sharp edges, areas of mostly one color, etc. \cite{Loy,Geron}.

\paragraph*{}The convolution layers allows for several advantages. Among these are (i) \textit{sparse-connectivity}: not every single activation (pixel) is connect to every single output pixel, so it is more computationally efficient, (ii) \textit{positional invariance}: key features can be identified regardless of where they are in the layer

\paragraph*{}A Maximum Pooling layers simply returns the maximum neuron activation in a group of neurons. In the case of 2D Max Pooling, we choose a kernel size to be $m \times n$, just like in the 2D Convolution layer, and extract the maximum value in each window, and then step along accrdoing to the chosen stride size \cite{Loy,Goodfellow}.As an example, consider again figure (\ref{fig-2DConvExample}). Using the $2 \times 2$ kernel on the $3 \times 4$ input, each box would then contain the maximum value of the input activations. We detail this in fig. (\ref{fig-2DMaxPool}):
\begin{figure}[h]
\label{fig-2DMaxPool}
\begin{center}
\includegraphics[scale=0.6]{../figures/2DConvExample(d)}
\end{center}
\caption{The result of 2D maximum-pooling an input array. This image was adapted from Loy, pg. 126 \cite{Loy}}
\end{figure}

\paragraph*{}The 1-Dimensional Concatenation Layer, also called 1D-Concat layer takes separate vectors of activations and combines them into a single 1D vector. Consider the layer activations

\paragraph*{}The softmax activation function is commonly used in the output layer of a multi-class classification network, such as the one we implement. When softmax acts on a vector or activation, the result is a normalized,non-negative output vector. \cite{Geron,Goodfellow,Virtanen}. The $i$-th element in a softmax activation function is given by:
\begin{equation}
\label{eqn-Softmax}
\text{Softmax}\big[ x \big]_{i} = \frac{1}{\sum_{j}e^{x_j}}e^{x_i}
\end{equation}

\paragraph*{}This function is only used in the output layer of a neural network, and encodes a discrete categorical probability. For example, if eqn.(\ref{eqn-Softmax}) returns a vector of the form: $[ 0.75 , 0.25]$, we interpret this as a sample having a $75\%$ chance of beloinging to class $0$ and a $25\%$ change of belonging to class $1$.

\begin{algorithmic}
\REQUIRE Cost/Objective function $J$ and learning rate $\alpha$
\REQUIRE Set of weighting parameters - $W^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of bias parameters - $b^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of layer activation function - $a^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of layer activation function derivatives - $\partial a^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of pre-nonlinear activation - $Z^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of post-nonlinear activation - $X^{(i)}, i \in \{0,1,....,L-1\}$

\FOR {$L-1,L-2,...,2,1$}
	\item Compute gradient w.r.t pre-nonlinear activation portion of layer function
	\item $dx^{(l)} \leftarrow \nabla_{Z^{(l)}}J = dx^{(l)} \odot \partial a^{(l)}[ Z^{(l)} ]$
	\item Compute gradient w.r.t weighting and bias elements
	\item $db \leftarrow \nabla_{b^{(l)}}J = dx^{(l)}$
	\item $dW \leftarrow \nabla_{W^{(l)}}J = dx^{(l)} \cdot X^{(l-1)}$

@@

\item[•]Nathan Richard for consultation on physics, mathematics, and programming topics.

\item[•]Morgan Saidel for consultation on physics, mathematics, and programming topics.

@@

\item[•]John Parker for consultation on physics, mathematics, and programming topics.

\item[•]Dakota Buell for consultation on physics, mathematics, and programming topics.

\item[•]Dr. Kevin Short for various consultation and guidence throughout this project

\item[•]Dr. Maurik Holtrop for various consultation and guidence throughout this project

\item[•]University of Iowa, Electronic Music Studio for a digital sound library ued for training samples

\item[•]Philharmonics Symphony Orchestra for a digital sound library used for training samples

@@

>>>>

\paragraph*{}In secs. (\ref{subsec-NerualNetworkIntro}) and (\ref{subsec-NetworkStructure}), we introduce the nature and structure of a neural network and how it can use used in part to solve this problem. We show it's mathematical function analogue and how

@@

\subsection{The Structure of a Neural Network}
\label{subsec-NetworkStructure}

\begin{algorithmic}
\REQUIRE Network with $L$ layers - $0,1,2,....,L-2,L-1$
\REQUIRE Set of layer functions $f^{(l)}, l \in 0,1,2,....,L-2,L-1$ made up of a linear transformation and a non-linear transformation.
\REQUIRE Input sample - $x^{(0)}$

If training the network, the intermediate layer activations must be stored for back-propagation. Otherwise, they must be discarded.
$Z = \{\} \leftarrow$ array to hold pre-activations \\
$X = \{\} \leftarrow$ array to hold final activations \\

\FOR {$l = 1,2,...,L-2,L-1$}	
	\item Evaluate Linear Transformation
	\item $z^{(l)} \leftarrow$ 
	\item Apply non-linear Activation Function
	\item $x^{(l)} \leftarrow \sigma^{(l)} \big[ z^{(l)} \big]$ 
	\item Store values for later use in optimizer, otherwise can be discarded.
	\item $Z \leftarrow Z.add(z^{(l)})$
	\item $X \leftarrow X.add(x^{(l)})$
\ENDFOR

\paragraph*{}As stated previously, a neural network is composed of a series of functions that are called successively to transform features (an input) into a prediction (an output). As shown in eqn. (\ref{eqn-altLayerFunction}), each function feeds directly into the next as to form a sort of recursive computational graph \cite{Goodfellow}.

\paragraph*{}Typically, a layer function can be divided into two portions: (i.) a Linear transformation, with a bias addition, and (ii.) an element-wise activation transformation. This activation function typically is a non-linear function which allows for the modeling of increasingly complex decision-boundaries. The Linear transformation can be given by eqn (\ref{eqn-LinearTransform}) in the case of a 1D Dense layer (\ref{subsubsec-DenseLayer}) or eqn. (\ref{eqn-LinearTransform2}) in the case of a 2D Convolution layer.
\begin{equation}
\label{eqn-LinearTransform}
z^{(l)} = W^{(l)} x^{(l-1)} + b^{(l)}
\end{equation}
\begin{equation}
\label{eqn-LinearTransform2}
z^{(l)} = \big(W^{(l)} * x^{(l-1)}\big) + b^{(l)}
\end{equation} 
Where $W^{(l)} x^{(l-1)}$ denotes a matrix-vector product, and $\big(W^{(l)} * x^{(l-1)}\big)$ denotes a convolution product. In both cases, $W^{(l)}$ is the \textit{weighting-matrix} or \textit{weighting-kernel} for layer $l$, $b^{(l)}$ is the \textit{bias-vector} for layer $l$, $z^{(l)}$ are the \textit{linear-activations} for layer $l$ and $x^{(l-1)}$ is the final activations for layer $l-1$.

\paragraph*{}Step (ii.) is usually given by some \textit{activation function}:
\begin{equation}
\label{eqn-elementActivation}
x^{(l)} = \sigma^{(l)} \big[ z^{(l)} \big]
\end{equation}
Where $x^{(l)}$ is the final activations for layer $l$ and $z^{(l)}$ is given in equation (\ref{eqn-LinearTransform}). $\sigma^{(l)}$ is some activation function, which is often use to enable the modeling of more complex-decision boundaries.

\paragraph*{}In the case of a 2D input array, $x^{(l-1)} \in \mathbb{R}^{(N \times M)}$, the convolutions filters "step" through the input data and repeatedly compute the element-wise product of each $m \times n$ weighting kernel, and the local activation of the $x^{(l-1)}$ array. The step size in each of the 2-dimension is known as the \textit{stride}. Each of the $K$ filters are used to generate a $m \times n$ feature map from the input activation. For an appropriately trained network, some filters may be dedicated to detecting vertical lines, horizontal lines, sharp edges, areas of mostly one color, etc. \cite{Geron,Loy}.

\paragraph*{}The convolution layers allows for several advantages. Among these are (i) \textit{sparse-connectivity}: not every single activation (pixel) is connect to every single output pixel, so it is more computationally efficient, (ii) \textit{positional invariance}: key features can be identified regardless of where they are in the layer, and (iii.) \textit{Automatic feature detection} as the training process will update the filters to identify dominant features in the data without human instruction \cite{Geron,Goodfellow,Loy}.

\paragraph*{}A Maximum Pooling layers simply returns the maximum neuron activation in a group of neurons. In the case of 2D Max Pooling, we choose a kernel size to be $m \times n$, just like in the 2D Convolution layer, and extract the maximum value in each window, and then step along accrdoing to the chosen stride size \cite{Loy,Goodfellow}.As an example, consider again figure (\ref{fig-2DConvExample}a). Using the $2 \times 2$ kernel on the $3 \times 4$ input, each box would then contain the maximum value of the input activations. We detail this in fig. (\ref{fig-2DMaxPool}):
\begin{figure}[h]
\label{fig-2DMaxPool}
\begin{center}
\includegraphics[scale=0.6]{../figures/2DConvExample(d)}
\end{center}
\caption{The result of 2D maximum-pooling an input array. This image was adapted from Loy, pg. 126 \cite{Loy}}
\end{figure}

\paragraph*{}The 1-Dimensional Concatenation Layer, also called 1D-Concat layer takes separate vectors of activations and combines them into a single 1D vector. Consider the layer activations $\vec{a}^{(l-1)}$ and $\vec{b}^{(l-1)}$ with shapes  $1 \times \alpha$ and $1 \times \beta$ respectively. They are the outputs of two different layers:
\begin{equation}
\vec{a}^{(l-1)} = \big[ a_{0},  a_{1}, ... ,  a_{\alpha-1} \big]
\text{\hspace{1cm} and \hspace{1cm}}
\vec{b}^{(l-1)} = \big[ b_{0},  b_{1}, ... ,  b_{\beta-1} \big]
\end{equation}
The result of the concatenation is a new 1D-array, $\vec{c}^{(l)}$ with size $1 \times \alpha\beta$:
\begin{equation}
\vec{c}^{(l)} = \big[ a_{0},  a_{1}, ... ,  a_{\alpha-1}, b_{0},  b_{1}, ... ,  b_{\beta-1} \big]
\end{equation}

\paragraph*{}We can denote this for $n$ arrays with function notation:
\begin{equation}
\label{eqn-ConcatenationFunction}
f^{(l)} : x^{(l-1)}_{a} \in \mathbb{R}^{(1 \times \alpha)} , x^{(l-1)}_{b} \in \mathbb{R}^{(1 \times \beta)} , ....  \rightarrow
x^{(l)}_{z} \in \mathbb{R}^{(1 \times \alpha\beta...)}
\end{equation}
The concatenation layer is used to combine activation non-adjacent, or non-related layers into a single new layer. In this case of the model used in the project, we use it to combine the activations that result from two distinct models into a single layer that is then transformed into a final output. This process is detailed further in sec. (\ref{sec-Architecture}).

\paragraph*{}The softmax activation function is commonly used in the output layer of a multi-class classification network, such as the one we implement. When softmax acts on a vector or activation, the result a non-negative output vector. with an $L_1$ norm of $1$ \cite{Geron,Goodfellow,Virtanen}. The $i$-th element in a softmax activation function is given by:
\begin{equation}
\label{eqn-Softmax}
\text{Softmax}\big[ x \big]_{i} = \frac{1}{\sum_{j}e^{x_j}}e^{x_i}
\end{equation}

\paragraph*{}This function is almost only used in the output layer of a neural network, and encodes a discrete categorical probability. For example, if eqn.(\ref{eqn-Softmax}) returns a vector of the form: $[ 0.75 , 0.25]$, we interpret this as a sample having a $75\%$ chance of beloinging to class $0$ and a $25\%$ change of belonging to class $1$.

\begin{algorithmic}
\REQUIRE Cost/Objective function $J$ and learning rate $\alpha$
\REQUIRE Set of weighting parameters - $W^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of bias parameters - $b^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of layer activation function - $\sigma^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of layer activation function derivatives - $\partial\sigma^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of pre-nonlinear activation - $Z^{(i)}, i \in \{0,1,....,L-1\}$
\REQUIRE Set of post-nonlinear activation - $X^{(i)}, i \in \{0,1,....,L-1\}$

\FOR {$L-1,L-2,...,2,1$}
	\item Compute gradient w.r.t pre-nonlinear activation portion of layer function
	\item $dx^{(l)} \leftarrow \nabla_{Z^{(l)}}J = dx^{(l)} \odot \partial\sigma^{(l)}[ Z^{(l)} ]$
	\item Compute gradient w.r.t weighting and bias elements
	\item $db \leftarrow \nabla_{b^{(l)}}J = dx^{(l)}$
	\item $dW \leftarrow \nabla_{W^{(l)}}J = dx^{(l)} \cdot X^{(l-1)}$

@@

@@

\item[•]Dakota Buell, John Parker, Nathan Richard, Morgan Saidel for consultation on physics, mathematics, and programming topics.

@@

\item[•]Dr. Kevin Short and Dr. Maurik Holtrop for various consultation and guidance throughout this project

\item[•]University of Iowa, Electronic Music Studio and Philharmonica Symphony Orchestra for the digital sound library used for training data samples.

@@

\bibitem{Tenorflow}
TensorFlow: Large-scale machine learning on heterogeneous systems,
2015. Software available from tensorflow.org.
